Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population

### 2024-04-24 analysis/paper/paper.tex
add extra files from post-rendering

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=30mm,bottom=30mm,left=30mm,heightrounded]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{lineno}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Fig}
\else
  \newcommand\figurename{Fig}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population},
  pdfauthor={Bjørn Peare Bartholdy; Jørgen B. Hasselstrøm; Lambert K. Sørensen; Maia Casna; Menno Hoogland; Historisch Genootschap Beemster; Amanda G. Henry},
  pdfkeywords={dental calculus; LC-MS/MS; alkaloids; dental pathology;
sinusitis; caffeine; tobacco},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

%\maketitle

@@

%\thispagestyle{fancy}

\LARGE{Multiproxy analysis exploring patterns of diet and disease in
dental calculus and skeletal remains from a 19th century Dutch
population}

@@

\normalsize{Bjørn Peare
Bartholdy}\textsuperscript{1,2*}, \normalsize{Jørgen B.
Hasselstrøm}\textsuperscript{3}, \normalsize{Lambert K.
Sørensen}\textsuperscript{3}, \normalsize{Maia
Casna}\textsuperscript{1}, \normalsize{Menno
Hoogland}\textsuperscript{1}, \normalsize{Historisch Genootschap
Beemster}\textsuperscript{4}, \normalsize{Amanda G.
Henry}\textsuperscript{1}\vspace{12pt}

%\rule{\textwidth}{1.8pt}\vspace{8pt}

@@

\textsuperscript{\footnotesize{1}}\textit{\small{Department of
Archaeological Sciences, Leiden
University}}\\ \textsuperscript{\footnotesize{2}}\textit{\small{Faculty
of Mechanical, Maritime and Materials Engineering, Delft University of
Technology}}\\ \textsuperscript{\footnotesize{3}}\textit{\small{Department
of Forensic Medicine, Aarhus
University}}\\ \textsuperscript{\footnotesize{4}}\textit{\small{Historisch
Genootschap Beemster}}

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* \small{b.p.bartholdy@tudelft.nl}

@@

\linenumbers

@@

%\vfill

%\newgeometry{top=30mm,left=30mm}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, interior hidden, boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

\hypertarget{materials}{%
\section{Materials}\label{materials}}

@@

For our sample, we preferentially selected males from the middle adult
age category (35-49 years) to minimise the effect of confounding
cultural and biological factors. Previous research on Middenbeemster has
shown a gendered division of labour (Palmer et al., 2016), and there are
biological differences in dental calculus formation and drug metabolism
that are related to age and sex (Huang et al., 2023; Uno et al., 2017;
White, 1997). The sample consists of 27 males, 11 probable males, 2
probable females, and 1 female (\textbf{?@fig-sample-demography}). We
selected males due to a higher occurrence of pipe notches and dental
calculus deposits than females (unpublished observation).

\emph{Fig 1 (fig-sample-demography): Overview of sample demography. Left
plot is the first batch and right plot is the replication batch with 29
of the individuals from the first batch. eya = early young adult (18-24
years); lya = late young adult (25-34 years); ma = middle adult (35-49
years); old = old adult (50+ years). Male? = probable male; Female? =
probable female.}

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{skeletal-analysis}{%
\subsection{Skeletal analysis}\label{skeletal-analysis}}

Demographic and pathological analyses were conducted in the Laboratory
for Human Osteoarchaeology at Leiden University. Sex was estimated using
cranial and pelvic morphological traits (Buikstra \& Ubelaker, 1994).
Age-at-death was estimated using dental wear, auricular and pubic
surface appearance, cranial suture closure, and epiphyseal fusion
(Brooks \& Suchey, 1990; Buckberry \& Chamberlain, 2002; Buikstra \&
Ubelaker, 1994; Lovejoy et al., 1985; Meindl \& Lovejoy, 1985), and
divided into the following categories: early young adult (18-24 years),
late young adult (25-34 years), middle adult ( 35-49 years), old adult
(50+ years). Preservation was visually scored on a four-stage scale
(excellent, good, fair, poor) based on the surface condition of the
bones and the extent of taphonomic degradation.

\hypertarget{paleopathology}{%
\subsubsection{Paleopathology}\label{paleopathology}}

@@

\hypertarget{dental-pathology}{%
\subsubsection{Dental pathology}\label{dental-pathology}}

@@

\hypertarget{calculus-sampling}{%
\subsection{Calculus sampling}\label{calculus-sampling}}

@@

\hypertarget{uhplc-msms}{%
\subsection{UHPLC-MS/MS}\label{uhplc-msms}}

@@

\hypertarget{statistical-analysis}{%
\subsection{Statistical analysis}\label{statistical-analysis}}

@@

All statistical analysis was conducted in R version 4.3.2 (2023-10-31),
Eye Holes, (R Core Team, 2020). Data wrangling was conducted with the
\textbf{tidyverse} (Wickham et al., 2019) and visualisations were
created using \textbf{ggplot2} (Wickham, 2016). Polychoric correlations
were calculated with the \textbf{psych} package (Revelle, 2022).

\hypertarget{results}{%
\section{Results}\label{results}}

@@

\hypertarget{tbl-compound-detect}{}
\begin{longtable}[]{@{}lllr@{}}
\caption{\label{tbl-compound-detect}Target compound including whether it
was detected (TRUE) or not (FALSE) in each batch, as well as the lower
limit of quantitation (LLOQ) in ng. CBD = cannabidiol; CBN = cannabinol;
THC = tetrahydrocannabinol; THCA-A = tetrahydrocannabinolic acid A;
THCVA = tetrahydrocannabivarin acid.}\tabularnewline
\toprule\noalign{}
Compound & Batch 1 & Batch 2 & LLOQ \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Compound & Batch 1 & Batch 2 & LLOQ \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
CBD & TRUE & FALSE & 0.050 \\
CBN & TRUE & FALSE & 0.050 \\
Caffeine & TRUE & TRUE & 0.050 \\
Cocaine & TRUE & FALSE & 0.025 \\
Cotinine & TRUE & TRUE & 0.050 \\
Nicotine & TRUE & TRUE & 0.100 \\
Salicylic acid & TRUE & TRUE & 0.500 \\
THC & TRUE & FALSE & 0.100 \\
THCA-A & TRUE & FALSE & 0.025 \\
THCVA & TRUE & FALSE & 0.010 \\
Theophylline & TRUE & TRUE & 0.010 \\
\end{longtable}

@@

Most plots show a large increase in extracted mass of a compound between
the calculus wash extracts (wash 1-3) and the dissolved calculus (calc).
Most samples containing theophylline and caffeine had the largest
quantity of the compound extracted from the first wash, then decreasing
in washes 2 and 3. There is an increase between wash 3 and the dissolved
calculus in all samples. The patterns are consistent across batches 1
and 2. Nicotine and cotinine have the same relative quantities in the
samples, i.e., the sample with the highest extracted quantity of
nicotine also had the highest extracted quantity of cotinine
(\textbf{?@fig-auth-plot-batch2}).

\emph{Fig 2 (fig-auth-plot-batch2): (A) Number of samples in which each
compound was detected in the first and second batch. (B) Quantity (ng)
of each compound extracted from each sample in batch 2. The plot
displays the extracted quantity across the three washes and final
calculus extraction (calc). Each coloured line represents a different
calculus sample. CBD = cannabidiol; CBN = cannabinol; THC =
tetrahydrocannabinol; THCA-A = tetrahydrocannabinolic acid A; THCVA =
tetrahydrocannabivarin acid.}

To see if preservation of the skeletal remains had any effect on the
detection of compounds, we compare extracted quantities of compounds to
the various levels of skeletal preservation. Our results from batch 2
suggest that detection of a compound may be linked to the preservation
of the skeleton, with better preservation leading to increased
extraction quantity (\textbf{?@fig-detection-preservation}A). We also
find a weak positive correlation between the weight of the calculus
sample and the quantity of compound extracted from the calculus
(\textbf{?@fig-detection-preservation}B).

\emph{Fig 3 (fig-detection-preservation): (A) Violin plot with overlaid
box plots depicting the distribution of extracted quantities of each
compound from batch 2 separated by state of preservation of the
skeleton. (B) Extracted quantity (ng) of compound plotted against
weights of the calculus samples from batch 2. r = Pearson correlation
coefficient.}

@@

\hypertarget{correlations-between-detected-alkaloids-and-diseases}{%
\subsection{Correlations between detected alkaloids and
diseases}\label{correlations-between-detected-alkaloids-and-diseases}}

@@

Polychoric correlation was conducted on the dichotomised compounds and
pathological conditions, as well as the discretised dental diseases.
Salicylic acid was removed due to its ubiquitous presence in the sample,
and is likely to cause spurious correlations. Strong correlations were
found between cotinine and nicotine (0.85). Moderate correlations were
found between OA and DDD (0.47), VOP and periodontitis (0.49), SN and
cotinine (0.56), DDD and calculus (-0.42), CMS and caffeine (0.53),
caries and periodontitis (0.52), periodontitis and VOP (0.49),
periodontitis and age-at-death (0.41), nicotine and SN (0.53), calculus
and DDD (-0.42), age-at-death and theophylline (-0.45), theophylline and
age-at-death (-0.45), caffeine and periodontitis (0.49), cotinine and
CMS (0.43). Remaining correlations were weak or absent
(\textbf{?@fig-polycorr}). Correlations with age will be depressed
because age was largely controlled for in the sample selection.

\emph{Fig 4 (fig-polycorr): Plot of the polychoric correlations
(}rho\emph{). Larger circles and increased opacity indicates a stronger
correlation coefficient. OA = osteoarthritis; VOP = vertebral
osteophytosis; SN = Schmorl's nodes; DDD = degenerative disc disease; CO
= cribra orbitalia; CMS = chronic maxillary sinusitis; SA = salicylic
acid.}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

@@

Nicotine and its principal/main metabolite, cotinine, were strongly
positively correlated, both in concentration and presence/absence in
individuals (Table~\ref{tbl-pearson} and \textbf{?@fig-polycorr}). The
detection of nicotine and cotinine is not surprising, as pipe-smoking in
the Beemsterpolder is well-documented in the literature (Aten et al.,
2012; Bouman, 2017), and visible on the skeletal remains as pipe notches
(Lemmers et al., 2013). There is also documented medicinal use of
nicotine in the Beemsterpolder, where a tobacco-smoke enema was used for
headaches, respiratory problems, colds, and drowsiness from around 1780
to 1830 (Aten et al., 2012). In our sample, we also detected nicotine
and cotinine (replicated) in an old adult, probable female individual.
In this particular case it is unlikely that the compounds entered the
dental calculus through pipe-smoking, as the individual had no visible
pipe notches; more likely the tobacco entered through an alternate mode
of consumption, secondhand smoke, or the aforementioned tobacco-smoke
enema.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=10mm,bottom=10mm,left=30mm,heightrounded]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{fancyhdr, graphicx}
\usepackage[hidelinks]{hyperref}

\fancyhead{\vspace{70pt}}
\renewcommand{\headrulewidth}{1.8pt}

\makeatletter
\def\@maketitle{%
  \newpage
  \hfill\href{https://doi.org/10.24072/pci.archaeo.100389}{\includegraphics[width=3.5cm]{badge_PCI_Archaeology.png}}\hfill
  \null
 \begin{center}
{\LARGE
\textbf\newline{\@title}
}
%\newline

\end{center}
  \par}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population},
  pdfauthor={Bjørn Peare Bartholdy; Jørgen B. Hasselstrøm; Lambert K. Sørensen; Maia Casna; Menno Hoogland; Historisch Genootschap Beemster; Amanda G. Henry},
  pdfkeywords={dental calculus; LC-MS/MS; alkaloids; dental pathology;
sinusitis; caffeine; tobacco},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\maketitle

@@

\thispagestyle{fancy}

@@

\normalsize{Bjørn Peare
Bartholdy}\textsuperscript{1*}, \normalsize{Jørgen B.
Hasselstrøm}\textsuperscript{2}, \normalsize{Lambert K.
Sørensen}\textsuperscript{2}, \normalsize{Maia
Casna}\textsuperscript{1}, \normalsize{Menno
Hoogland}\textsuperscript{1}, \normalsize{Historisch Genootschap
Beemster}\textsuperscript{3}, \normalsize{Amanda G.
Henry}\textsuperscript{1}\vspace{12pt}

\rule{\textwidth}{1.8pt}\vspace{8pt}

@@

\textsuperscript{\footnotesize{1}}\textit{\small{Department of
Archaeological Sciences, Leiden
University}}\\ \textsuperscript{\footnotesize{2}}\textit{\small{Department
of Forensic Medicine, Aarhus
University}}\\ \textsuperscript{\footnotesize{3}}\textit{\small{Historisch
Genootschap Beemster}}

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* \small{b.p.bartholdy@arch.leidenuniv.nl}

@@

@@

\vfill

\newgeometry{top=30mm,left=30mm}
\section{Introduction}\label{introduction}

@@

\section{Materials}\label{materials}

@@

For our sample, we preferentially selected males from the middle adult
age category (35-49 years) to minimise the effect of confounding
cultural and biological factors. Previous research on Middenbeemster has
shown a gendered division of labour (Palmer et al., 2016), and there are
biological differences in dental calculus formation and drug metabolism
that are related to age and sex (Huang et al., 2023; Uno et al., 2017;
White, 1997). The sample consists of 27 males, 11 probable males, 2
probable females, and 1 female (Figure~\ref{fig-sample-demography}). We
selected males due to a higher occurrence of pipe notches and dental
calculus deposits than females (unpublished observation).

\begin{figure}

\centering{

\includegraphics{../figures/fig-sample-demography-1.pdf}

}

\caption{\label{fig-sample-demography}Overview of sample demography.
Left plot is the first batch and right plot is the replication batch
with 29 of the individuals from the first batch. eya = early young adult
(18-24 years); lya = late young adult (25-34 years); ma = middle adult
(35-49 years); old = old adult (50+ years). Male? = probable male;
Female? = probable female.}

\end{figure}%

\section{Methods}\label{methods}

\subsection{Skeletal analysis}\label{skeletal-analysis}

Demographic and pathological analyses were conducted in the Laboratory
for Human Osteoarchaeology at Leiden University. Sex was estimated using
cranial and pelvic morphological traits (Buikstra \& Ubelaker, 1994).
Age-at-death was estimated using dental wear, auricular and pubic
surface appearance, cranial suture closure, and epiphyseal fusion
(Brooks \& Suchey, 1990; Buckberry \& Chamberlain, 2002; Buikstra \&
Ubelaker, 1994; Lovejoy et al., 1985; Meindl \& Lovejoy, 1985), and
divided into the following categories: early young adult (18-24 years),
late young adult (25-34 years), middle adult (35-49 years), old adult
(50+ years). Preservation was visually scored on a four-stage scale
(excellent, good, fair, poor) based on the surface condition of the
bones and the extent of taphonomic degradation.

\subsubsection{Paleopathology}\label{paleopathology}

@@

\subsubsection{Dental pathology}\label{dental-pathology}

@@

\subsection{Calculus sampling}\label{calculus-sampling}

@@

\subsection{UHPLC-MS/MS}\label{uhplc-msms}

@@

\subsection{Statistical analysis}\label{statistical-analysis}

@@

All statistical analysis was conducted in R version 4.3.3 (2024-02-29),
Angel Food Cake, (R Core Team, 2020). Data wrangling was conducted with
the \textbf{tidyverse} (Wickham et al., 2019) and visualisations were
created using \textbf{ggplot2} (Wickham, 2016). Polychoric correlations
were calculated with the \textbf{psych} package (Revelle, 2022).

\section{Results}\label{results}

@@

\begin{longtable}[]{@{}lllr@{}}

\caption{\label{tbl-compound-detect}Target compound including whether it
was detected (TRUE) or not (FALSE) in each batch, as well as the lower
limit of quantitation (LLOQ) in ng. CBD = cannabidiol; CBN = cannabinol;
THC = tetrahydrocannabinol; THCA-A = tetrahydrocannabinolic acid A;
THCVA = tetrahydrocannabivarin acid.}

\tabularnewline

\toprule\noalign{}
Compound & Batch 1 & Batch 2 & LLOQ \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
CBD & TRUE & FALSE & 0.050 \\
CBN & TRUE & FALSE & 0.050 \\
Caffeine & TRUE & TRUE & 0.050 \\
Cocaine & TRUE & FALSE & 0.025 \\
Cotinine & TRUE & TRUE & 0.050 \\
Nicotine & TRUE & TRUE & 0.100 \\
Salicylic acid & TRUE & TRUE & 0.500 \\
THC & TRUE & FALSE & 0.100 \\
THCA-A & TRUE & FALSE & 0.025 \\
THCVA & TRUE & FALSE & 0.010 \\
Theophylline & TRUE & TRUE & 0.010 \\

\end{longtable}

@@

Most plots show a large increase in extracted mass of a compound between
the calculus wash extracts (wash 1-3) and the dissolved calculus (calc).
Most samples containing theophylline and caffeine had the largest
quantity of the compound extracted from the first wash, then decreasing
in washes 2 and 3. There is an increase between wash 3 and the dissolved
calculus in all samples. The patterns are consistent across batches 1
and 2. Nicotine and cotinine have the same relative quantities in the
samples, i.e., the sample with the highest extracted quantity of
nicotine also had the highest extracted quantity of cotinine
(Figure~\ref{fig-auth-plot-batch2}).

\begin{figure}

\centering{

\includegraphics{../figures/fig-auth-plot-batch2-1.pdf}

}

\caption{\label{fig-auth-plot-batch2}(A) Number of samples in which each
compound was detected in the first and second batch. (B) Quantity (ng)
of each compound extracted from each sample in batch 2. The plot
displays the extracted quantity across the three washes and final
calculus extraction (calc). Each coloured line represents a different
calculus sample. CBD = cannabidiol; CBN = cannabinol; THC =
tetrahydrocannabinol; THCA-A = tetrahydrocannabinolic acid A; THCVA =
tetrahydrocannabivarin acid.}

\end{figure}%

To see if preservation of the skeletal remains had any effect on the
detection of compounds, we compare extracted quantities of compounds to
the various levels of skeletal preservation. Our results from batch 2
suggest that detection of a compound may be linked to the preservation
of the skeleton, with better preservation leading to increased
extraction quantity (Figure~\ref{fig-detection-preservation}A). We also
find a weak positive correlation between the weight of the calculus
sample and the quantity of compound extracted from the calculus
(Figure~\ref{fig-detection-preservation}B).

\begin{figure}

\centering{

\includegraphics{../figures/fig-detection-preservation-1.pdf}

}

\caption{\label{fig-detection-preservation}(A) Violin plot with overlaid
box plots depicting the distribution of extracted quantities of each
compound from batch 2 separated by state of preservation of the
skeleton. (B) Extracted quantity (ng) of compound plotted against
weights of the calculus samples from batch 2. r = Pearson correlation
coefficient.}

\end{figure}%

@@

\subsection{Correlations between detected alkaloids and
diseases}\label{correlations-between-detected-alkaloids-and-diseases}

@@

Polychoric correlation was conducted on the dichotomised compounds and
pathological conditions, as well as the discretised dental diseases.
Salicylic acid was removed due to its ubiquitous presence in the sample,
and is likely to cause spurious correlations. Strong correlations were
found between cotinine and nicotine (0.85). Moderate correlations were
found between OA and DDD (0.47), VOP and periodontitis (0.49), SN and
cotinine (0.56), DDD and calculus (-0.42), CMS and caffeine (0.53),
caries and periodontitis (0.52), periodontitis and VOP (0.49),
periodontitis and age-at-death (0.41), nicotine and SN (0.53), calculus
and DDD (-0.42), age-at-death and theophylline (-0.45), theophylline and
age-at-death (-0.45), caffeine and periodontitis (0.49), cotinine and
CMS (0.43). Remaining correlations were weak or absent
(Figure~\ref{fig-polycorr}). Correlations with age will be depressed
because age was largely controlled for in the sample selection.

\begin{figure}

\centering{

\includegraphics{../figures/fig-polycorr-1.pdf}

}

\caption{\label{fig-polycorr}Plot of the polychoric correlations
(\emph{rho}). Larger circles and increased opacity indicates a stronger
correlation coefficient. OA = osteoarthritis; VOP = vertebral
osteophytosis; SN = Schmorl's nodes; DDD = degenerative disc disease; CO
= cribra orbitalia; CMS = chronic maxillary sinusitis; SA = salicylic
acid.}

\end{figure}%

\section{Discussion}\label{discussion}

@@

Nicotine and its principal/main metabolite, cotinine, were strongly
positively correlated, both in concentration and presence/absence in
individuals (Table~\ref{tbl-pearson} and Figure~\ref{fig-polycorr}). The
detection of nicotine and cotinine is not surprising, as pipe-smoking in
the Beemsterpolder is well-documented in the literature (Aten et al.,
2012; Bouman, 2017), and visible on the skeletal remains as pipe notches
(Lemmers et al., 2013). There is also documented medicinal use of
nicotine in the Beemsterpolder, where a tobacco-smoke enema was used for
headaches, respiratory problems, colds, and drowsiness from around 1780
to 1830 (Aten et al., 2012). In our sample, we also detected nicotine
and cotinine (replicated) in an old adult, probable female individual.
In this particular case it is unlikely that the compounds entered the
dental calculus through pipe-smoking, as the individual had no visible
pipe notches; more likely the tobacco entered through an alternate mode
of consumption, secondhand smoke, or the aforementioned tobacco-smoke
enema.
